### 2025-07-21 paper/quarto_aaai/paper.tex
editorial pass until section 3.4

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. The authors propose a way
to generate counterfactuals that are maximally faithful to the model in
that they are consistent with what the model has learned about the
underlying data. Formally, they rely on tools from energy-based
modelling \citep{teh2003energy} to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their counterfactual explainer,
\emph{ECCCo}, yields plausible explanations if and only if the
underlying model has learned representations that align with them. The
authors find that both deep ensembles \citep{lakshminarayanan2016simple}
and joint energy-based models (JEMs) \citep{grathwohl2020your}, a form
of constrastive learning, tend to do well in this regard.

It helps to look at these findings through the lens of representation
learning with high degrees of freedom. Deep ensembles are approximate
Bayesian model averages, which are particularly effective when models
are underspecified by the available data \citep{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of
overrelying on a single locally optimal representation that corresponds
to semantically meaningless explanations. Likewise, previous work of
\citet{schut2021generating} found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
undergone adversarial training. The case for JEMs is even clearer: they
optimize a hybrid objective that induces both high predictive
performance and strong generative capacity \citep{grathwohl2020your},
which resembles the idea of aligning models with plausible explanations
and has inspired CT.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values to ensure that
individuals seeking recourse through CE face minimal costs in terms of
feature changes. Different variations of Equation~\ref{eq-general} have
been proposed in the literature to address many desiderata including the
ones discussed above (faithfulness, plausibility and actionability).
Like \citet{wachter2017counterfactual}, most of these approaches rely on
gradient descent to optimize Equation~\ref{eq-general}. For more details
on the approaches tested in this work, we refer the reader to the
supplementary appendix. In the following, we describe in detail how
counterfactuals are generated and used in counterfactual training.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:

\begin{itemize}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
\end{itemize}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about the input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
choose to only consider domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
classification tasks for reasons discussed in
Section~\ref{sec-discussion}.

@@

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-offs between these components are
adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\). The full training regime is sketched out in
Algorithm \ref{alg-experiment}.

@@

\citet{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model that learns to discriminate
output classes conditional on the observed (training) samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood:
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
where the first term can be optimized using cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
minimize the contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
perform well on the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
(SGLD) with an uninformative prior for initialization but we depart from
their methodology. Instead we propose to leverage counterfactual
explainers to generate counterfactuals of observed training samples.
Specifically, we have: \begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while increasing the
energy of counterfactuals (negative samples) \citep{du2019implicit}. As
the counterfactuals get more plausible (Def. \ref{def-explainability})
during training, these opposing effects gradually balance each other out
\citep{lippe2024uvadlc}.

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods support domain and mutability
constraints. In principle, any existing approach for generating CEs is
viable, so long as it does not violate the faithfulness condition. Like
JEMs \citep{murphy2022probabilistic}, counterfactual training can be
considered a form of contrastive representation learning.

>>>>

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. The authors propose a way
to generate counterfactuals that are maximally faithful models in that
they are consistent with what models have learned about the underlying
data. Formally, they rely on tools from energy-based modelling
\citep{teh2003energy} to minimize the contrastive divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by a model. Their counterfactual explainer, \emph{ECCCo},
yields plausible explanations if and only if the underlying model has
learned representations that align with them. The authors find that both
deep ensembles \citep{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \citep{grathwohl2020your}, a form of constrastive
learning, tend to do well in this regard.

It helps to look at these findings through the lens of representation
learning with high degrees of freedom. Deep ensembles are approximate
Bayesian model averages, which are particularly effective when models
are underspecified by the available data \citep{wilson2020case}.
Averaging across solutions mitigates the risk of overrelying on a single
locally optimal representation that corresponds to semantically
meaningless explanations. Likewise, previous work of
\citet{schut2021generating} found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
undergone adversarial training. The case for JEMs is even clearer: they
optimize a hybrid objective that induces both high predictive
performance and strong generative capacity \citep{grathwohl2020your},
which resembles the idea of aligning models with plausible explanations
and has inspired CT.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\in\mathcal{Y}\) denotes some target
class. The \(\text{yloss}(\cdot)\) function quantifies the discrepancy
between current model predictions for \(\mathbf{x}^{\prime}\) and the
target class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values to ensure that
individuals seeking recourse through CE face minimal costs in terms of
feature changes. Different variations of Equation~\ref{eq-general} have
been proposed in the literature to address many desiderata including the
ones discussed above (faithfulness, plausibility and actionability).
Like \citet{wachter2017counterfactual}, most of these approaches rely on
gradient descent to optimize Equation~\ref{eq-general}. For more details
on the approaches tested in this work, we refer the reader to the
supplementary appendix. In the following, we describe in detail how
counterfactuals are generated and used in counterfactual training.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as follows:

\begin{itemize}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
\end{itemize}
Here, $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
adapted notation. Intuitively, plausible counterfactuals are consistent
with the data and faithful counterfactuals are consistent with what the
model has learned about the input data. Ac tionability constraints in
Def. \ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider mutability constraints for individual features \(x_d\) for
\(d=1,...,D\). We also limit ourselves to classification tasks for
reasons discussed in Section~\ref{sec-discussion}.

@@

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models,
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-offs between these components are
adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\). The full training regime is sketched out in
Algorithm \ref{alg-experiment}.

@@

\citet{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model that learns to discriminate
output classes conditional on the observed (training) samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood:
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
where the first term can be optimized using cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
minimize the contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and samples generated from \(p_\theta(\mathbf{x})\).

To generate samples, \citet{grathwohl2020your} use Stochastic Gradient
Langevin Dynamics (SGLD) with an uninformative prior for initialization
but we depart from their methodology: we propose to leverage
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:
\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Intuitively, the gradient of
Equation~\ref{eq-div} decreases the energy of observed training samples
(positive samples) while increasing the energy of counterfactuals
(negative samples) \citep{du2019implicit}. As the counterfactuals get
more plausible (Def. \ref{def-explainability}) during training, these
opposing effects gradually balance each other out
\citep{lippe2024uvadlc}.

Since maturity of counterfactuals in terms of a probability threshold is
often reached before \(T\), this form of sampling is not only more
closely aligned with Def. \ref{def-explainability}., but can also speed
up training times compared to SGLD. The departure from SGLD also allows
us to tap into the vast repertoire of explainers that have been proposed
in the literature to meet different desiderata. For example, many
methods support domain and mutability constraints. In principle, any
existing approach for generating CEs is viable, so long as it does not
violate the faithfulness condition. Like JEMs
\citep{murphy2022probabilistic}, counterfactual training can be
considered a form of contrastive representation learning.

### 2025-07-21 paper/sections/method.qmd
adjusted Algorithm 1 and ensured consistet capitalization for CT, AR, CE etc.
editorial pass until section 3.4

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address many desiderata including the ones discussed above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on the approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in counterfactual training.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:

\begin{itemize}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
\end{itemize}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology. Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:
$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (Def. \ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods support domain and mutability constraints. In principle, any existing approach for generating CEs is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], Counterfactual Training can be considered a form of contrastive representation learning.

>>>>

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+\in\mathcal{Y}$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address many desiderata including the ones discussed above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on the approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in counterfactual training.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as follows:

\begin{itemize}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
\end{itemize}
Here, $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} follows @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Ac tionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models,
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initialize model $\mathbf{M}_{\theta}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and samples generated from $p_\theta(\mathbf{x})$.

To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology: we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:
$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (Def. \ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

Since maturity of counterfactuals in terms of a probability threshold is often reached before $T$, this form of sampling is not only more closely aligned with Def. \ref{def-explainability}., but can also speed up training times compared to SGLD. The departure from SGLD also allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods support domain and mutability constraints. In principle, any existing approach for generating CEs is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], counterfactual training can be considered a form of contrastive representation learning.
